---
title: "RMS Titanic"
author: "Dakota Chang"
date: 2025-01-30
output:
  github_document:
    toc: true
prerequisites:
  - e-vis01-bar-charts
---

*Purpose*: Most datasets have at least a few variables. Part of our task in analyzing a dataset is to understand trends as they vary across these different variables. Unless we're careful and thorough, we can easily miss these patterns. In this challenge you'll analyze a dataset with a small number of categorical variables and try to find differences among the groups.

*Reading*: (Optional) [Wikipedia article](https://en.wikipedia.org/wiki/RMS_Titanic) on the RMS Titanic.

<!-- include-rubric -->
# Grading Rubric
<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual
<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|----------|----------------|--------------|
| Effort | Some task __q__'s left unattempted | All task __q__'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Submission
<!-- ------------------------- -->

Make sure to commit both the challenge report (`report.md` file) and supporting files (`report_files/` folder) when you are done! Then submit a link to Canvas. **Your Challenge submission is not complete without all files uploaded to GitHub.**


```{r setup}
library(tidyverse)

df_titanic <- as_tibble(Titanic)
```

*Background*: The RMS Titanic sank on its maiden voyage in 1912; about 67% of its passengers died.

# First Look
<!-- -------------------------------------------------- -->

### __q1__ Perform a glimpse of `df_titanic`. What variables are in this dataset?

```{r q1-task}
## TASK: Perform a `glimpse` of df_titanic
glimpse(df_titanic)
```


```{r}
## TASK: Perform a `glimpse` of df_titanic
df_titanic
```

```{r}
# More exploration
df_titanic$n %>% range()

df_titanic %>%
  select(where(is.character)) %>%
  summarise(across(everything(), ~ paste(unique(.), collapse = ", ")))
```

**Observations**:

Variables

- Class (chr): 1st, 2nd, 3rd, Crew

- Sex (chr): 	Male, Female

- Age (chr): Child, Adult

- Survived (chr): Yes, No

- n (num) [# of survivors from each category]: 0 <= x <= 670

### __q2__ Skim the [Wikipedia article](https://en.wikipedia.org/wiki/RMS_Titanic) on the RMS Titanic, and look for a total count of souls aboard. Compare against the total computed below. Are there any differences? Are those differences large or small? What might account for those differences?

```{r q2-summary}
## NOTE: No need to edit! We'll cover how to
## do this calculation in a later exercise.
df_titanic %>% summarize(total = sum(n))
```

**Observations**:

- Write your observations here

  - There are 2240 people on board.

- Are there any differences?

  - The dataset has 2201 people, which means 39 people on board is not accounted for here.

- If yes, what might account for those differences?
  - Unwillingness to be part of this data set
  - Lost information through time

### __q3__ Create a plot showing the count of persons who *did* survive, along with aesthetics for `Class` and `Sex`. Document your observations below.

*Note*: There are many ways to do this.

```{r q3-task}
## TASK: Visualize counts against `Class` and `Sex`

df_titanic %>% filter(Survived == "Yes") %>% 
  ggplot(aes(x = Class, y = n, fill=Sex)) +
  geom_col(position="dodge")+
  theme(
    text=element_text(family="Palatino"),
    axis.text.x = element_text(angle = 0, hjust = 1),  # Rotate x-axis labels
    strip.text = element_text(size = 10, face = "bold") # Style facet titles
  ) +
  labs(
    title = "Survival Counts for Class and Sex",
    x = "Classes",
    y = "Survival Count"
  )
```

**Observations**:

Survival counts for male crew members and female 1st class passengers are the highest at almost 200 and around 140. In 3rd class, male and female survival counts are similar. It is important to note that this is survival count, and not survival rates of each class and sex. 

# Deeper Look
<!-- -------------------------------------------------- -->

Raw counts give us a sense of totals, but they are not as useful for understanding differences between groups. This is because the differences we see in counts could be due to either the relative size of the group OR differences in outcomes for those groups. To make comparisons between groups, we should also consider *proportions*.[1]

The following code computes proportions within each `Class, Sex, Age` group.

```{r compute-prop}
## NOTE: No need to edit! We'll cover how to
## do this calculation in a later exercise.
df_prop <-
  df_titanic %>%
  group_by(Class, Sex, Age) %>%
  mutate(
    Total = sum(n),
    Prop = n / Total
  ) %>%
  ungroup()
df_prop
```

### __q4__ Replicate your visual from q3, but display `Prop` in place of `n`. Document your observations, and note any new/different observations you make in comparison with q3. Is there anything *fishy* in your plot?

```{r q4-task}
df_prop %>% 
  filter(Survived == "Yes") %>% 
  ggplot() +
  geom_col(mapping = aes(x = Class, y = Prop, fill = Sex), position = "dodge")+
  theme(
    text=element_text(family="Palatino"),
    axis.text.x = element_text(angle = 0, hjust = 1),  # Rotate x-axis labels
    strip.text = element_text(size = 10, face = "bold") # Style facet titles
  ) +
  labs(
    title = "Survival Counts for Class and Sex",
    x = "Classes",
    y = "Survival Count"
  )
```

**Observations**:

- Write your observations here.

First class and second class both have a 100% survival rate. Third class has the lowest survival rates, although third class males are more likely to survive than males in crew. Females are more likely to survive regardless of tier (disregarding 1st and 2nd).

- Is there anything *fishy* going on in your plot?

In the graph we observe that there is a 100% survival rate for both 1st and 2nd class passengers, which we know is not true according to the dataset.

  

### __q5__ Create a plot showing the group-proportion of occupants who *did* survive, along with aesthetics for `Class`, `Sex`, *and* `Age`. Document your observations below.

*Hint*: Don't forget that you can use `facet_grid` to help consider additional variables!

```{r q5-task}
df_titanic %>%
  group_by(Class, Sex, Age) %>%
  mutate(
    Total = sum(n),
  ) %>% 
  ungroup() %>% 
  filter(Survived == "Yes") %>%
  mutate(
    Prop = n / Total
  ) %>%
  ggplot(aes(x = factor(Class), y = Prop, fill = Sex)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "Survival Proportions by Class, Sex, and Age",
    x = "Class",
    y = "Proportions"
  ) +
  facet_grid(~ Age) +  # Faceting by Age to see proportions by different age groups
  theme_linedraw() +
  theme(
    text = element_text(family = "Palatino"),
    axis.text.x = element_text(angle = 45, hjust = 1),  # Rotate x-axis labels
    strip.text = element_text(size = 10, face = "bold")  # Style facet titles
  )

```

**Observations**:

It seems the survival proportions for females are much higher than for males across the board. In both age groups (Adults and Children), the highest survival proportions are seen in 1st class, with a noticeable difference between females and males for the adults class. For adults, the survival proportion is high for females in 1st class, while males have a much lower survival rate in the same class. For children, the survival proportions are 100% in both 1st and 2nd class, which suggests that children are either prioritised or more adept at survival for this situation. The Crew class shows lower survival proportions for both sexes compared to the 1st class, but female survival rates are still higher than male survival rates. Also, it seems that there are no children crew member on the Titanic recorded in this dataset.


- If you saw something *fishy* in q4 above, use your new plot to explain the fishy-ness.
  /

# Notes
<!-- -------------------------------------------------- -->

[1] This is basically the same idea as [Dimensional Analysis](https://en.wikipedia.org/wiki/Dimensional_analysis); computing proportions is akin to non-dimensionalizing a quantity.
