---
title: "Gapminder"
author: "(Your name here)"
date: 2020-
output:
  github_document:
    toc: true
prerequisites:
  - e-vis03-boxplots
---

*Purpose*: Learning to do EDA well takes practice! In this challenge you'll further practice EDA by first completing a guided exploration, then by conducting your own investigation. This challenge will also give you a chance to use the wide variety of visual tools we've been learning.

<!-- include-rubric -->
# Grading Rubric
<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual
<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|----------|----------------|--------------|
| Effort | Some task __q__'s left unattempted | All task __q__'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Submission
<!-- ------------------------- -->

Make sure to commit both the challenge report (`report.md` file) and supporting files (`report_files/` folder) when you are done! Then submit a link to Canvas. **Your Challenge submission is not complete without all files uploaded to GitHub.**


```{r setup}
library(tidyverse)
library(gapminder)

```

*Background*: [Gapminder](https://www.gapminder.org/about-gapminder/) is an
independent organization that seeks to educate people about the state of the
world. They seek to counteract the worldview constructed by a hype-driven media
cycle, and promote a "fact-based worldview" by focusing on data. The dataset
we'll study in this challenge is from Gapminder.

# Guided EDA
<!-- -------------------------------------------------- -->

First, we'll go through a round of *guided EDA*. Try to pay attention to the high-level process we're going through---after this guided round you'll be responsible for doing another cycle of EDA on your own!

### __q0__ Perform your "first checks" on the dataset. What variables are in this
dataset?

```{r q0-task}
## TASK: Do your "first checks" here!
# glimpse(gapminder)
gapminder
```

**Observations**:

- Write all variable names here

country
continent
year
lifeExp
pop
gdpPercap

### __q1__ Determine the most and least recent years in the `gapminder` dataset.

*Hint*: Use the `pull()` function to get a vector out of a tibble. (Rather than the `$` notation of base R.)

```{r q1-task}
## TASK: Find the largest and smallest values of `year` in `gapminder`
year_max <- gapminder %>% pull(year) %>% max()
year_min <- gapminder %>% pull(year) %>% min()
```

Use the following test to check your work.

```{r q1-tests}
## NOTE: No need to change this
assertthat::assert_that(year_max %% 7 == 5)
assertthat::assert_that(year_max %% 3 == 0)
assertthat::assert_that(year_min %% 7 == 6)
assertthat::assert_that(year_min %% 3 == 2)

if (is_tibble(year_max)) {
  print("year_max is a tibble; try using `pull()` to get a vector")
  assertthat::assert_that(False)
}

print("Nice!")
```

### __q2__ Filter on years matching `year_min`, and make a plot of the GDP per capita against continent. Choose an appropriate `geom_` to visualize the data. What observations can you make?

You may encounter difficulties in visualizing these data; if so document your challenges and attempt to produce the most informative visual you can.

```{r q2-task}
## TASK: Create a visual of gdpPercap vs continent
gapminder_1952 <- gapminder %>% filter(year == year_min)

ggplot(gapminder_1952, aes(x = continent, y = gdpPercap)) +
  geom_boxplot() +
  labs(title = "GDP per Capita by Continent (1952)",
       x = "Continent",
       y = "GDP per Capita") +
  theme_minimal()

outlier <- gapminder %>% filter(year == year_min) %>% filter(gdpPercap >= 30000)
print(outlier)

gapminder_1952 <- gapminder %>% filter(year == year_min) %>% filter(gdpPercap <= 30000)

ggplot(gapminder_1952, aes(x = continent, y = gdpPercap)) +
  geom_boxplot() +
  labs(title = "GDP per Capita by Continent (1952)",
       x = "Continent",
       y = "GDP per Capita") +
  theme_minimal()

library(maps)

world_map <- map_data("world")

# Adjust country names to match between gapminder and world map data
gapminder_1952 <- gapminder_1952 %>%
  mutate(country = case_when(
    country == "United States" ~ "USA",
    country == "United Kingdom" ~ "UK",
    country == "Congo, Dem. Rep." ~ "Democratic Republic of the Congo",
    country == "Congo, Rep." ~ "Republic of Congo",
    country == "Korea, Dem. Rep." ~ "North Korea",
    country == "Korea, Rep." ~ "South Korea",
    country == "Russian Federation" ~ "Russia",  
    country == "Soviet Union" ~ "Russia",        
    TRUE ~ as.character(country)
  ))

# Merge gapminder data with world map data
map_data <- world_map %>%
  left_join(gapminder_1952, by = c("region" = "country"))

ggplot(map_data, aes(x = long, y = lat, group = group, fill = gdpPercap)) +
  geom_polygon(color = "white", size = 0.1) +
  scale_fill_gradient(low = "lightblue", high = "darkblue", na.value = "gray", name = "GDP per Capita") +
  labs(title = "GDP per Capita by Country (1952)",
       subtitle = "Data from Gapminder",
       caption = "Gray areas indicate missing data") +
  theme_minimal() +
  theme(axis.text = element_blank(),
        axis.title = element_blank(),
        axis.ticks = element_blank(),
        panel.grid = element_blank())
```

**Observations**:

- Plotting it as a map, I see that a high concentration of countries in Oceania, Europe, and North America had a very high GDP. This aligns with the box plot. We also see from the map view that we might be missing quite a few countries from this dataset. In particular, Russia might be a big game changer for Europe's GDP.

**Difficulties & Approaches**:

- Unedited graph of data was very hard to read because of one specific outlier from Asia. -> I removed the outlier from the visualization. I also printed the outlier to inspect which country it is from. The country is Kuwait. According to Wikipedia, "By 1952, the country became the largest oil exporter in the Persian Gulf. This massive growth attracted many foreign workers, especially from Palestine, Egypt, Iran, and India." However, Kuwait was still a British protectorate at that time. The country was not independent until 1961. So, the data is not very reliable. I removed the outlier from the visualization.

### __q3__ You should have found *at least* three outliers in q2 (but possibly many more!). Identify those outliers (figure out which countries they are).

```{r q3-task}
## TASK: Identify the outliers from q2

# Bring back outliers
gapminder_1952 <- gapminder %>% filter(year == year_min)

outliers_1952 <- gapminder_1952 %>%
  group_by(continent) %>%
  filter(gdpPercap > quantile(gdpPercap, 0.75) + 1.5 * IQR(gdpPercap))

outliers_1952
```

**Observations**:

- Identify the outlier countries from q2
  - Angola
  - Bahrain
  - Canada
  - Gabon
  - Kuwait
  - South Africa
  - Switzerland
  - United States	Americas
  - Venezuela

*Hint*: For the next task, it's helpful to know a ggplot trick we'll learn in an upcoming exercise: You can use the `data` argument inside any `geom_*` to modify the data that will be plotted *by that geom only*. For instance, you can use this trick to filter a set of points to label:

```{r layer-filter}
## NOTE: No need to edit, use ideas from this in q4 below
gapminder %>%
  filter(year == max(year)) %>%

  ggplot(aes(continent, lifeExp)) +
  geom_boxplot() +
  geom_point(
    data = . %>% filter(country %in% c("United Kingdom", "Japan", "Zambia")),
    mapping = aes(color = country),
    size = 2
  )
```

### __q4__ Create a plot similar to yours from q2 studying both `year_min` and `year_max`. Find a way to highlight the outliers from q3 on your plot *in a way that lets you identify which country is which*. Compare the patterns between `year_min` and `year_max`.

*Hint*: We've learned a lot of different ways to show multiple variables; think about using different aesthetics or facets.

```{r q4-task}
## TASK: Create a visual of gdpPercap vs continent
```

**Observations**:

- Write your observations here

# Your Own EDA
<!-- -------------------------------------------------- -->

Now it's your turn! We just went through guided EDA considering the GDP per capita at two time points. You can continue looking at outliers, consider different years, repeat the exercise with `lifeExp`, consider the relationship between variables, or something else entirely.

### __q5__ Create *at least* three new figures below. With each figure, try to pose new questions about the data.

```{r q5-task1}
gapminder_2007 <- gapminder %>% filter(year == year_max)

# Plot life expectancy vs GDP per capita
ggplot(gapminder_2007, aes(x = gdpPercap, y = lifeExp)) +
  geom_point(aes(color = continent), size = 3, alpha = 0.7) +
  labs(title = "Life Expectancy vs GDP per Capita (2007)",
       x = "GDP per Capita",
       y = "Life Expectancy",
       color = "Continent") +
  theme_minimal()
    
```

- It seems that there is an exponential return for life expectancy as GDP per capita increases. However, it seems that there is a diminishing return at gdp of around 20,000. After that point, the life expectancy does not increase as much with increasing gdp. This is especially true for countries in Europe and North America. It also seems that many of the continents have a similar life expectancy. It is also important to note that except one country in Asia, all reported countries with a life expectancy of below 55 are in Africa.


```{r q5-task2}
## TASK: Your second graph
gapminder_pop <- gapminder %>%
  group_by(continent, year) %>%
  summarize(total_pop = sum(pop), .groups = 'drop')

# Plot population growth over time
ggplot(gapminder_pop, aes(x = year, y = total_pop, color = continent)) +
  geom_line(size = 1.2) +
  labs(title = "Population Growth Over Time by Continent",
       x = "Year",
       y = "Total Population",
       color = "Continent") +
  theme_minimal()
```

- While almost all countries have some form of growth in terms of population, Oceania and Europe had the least growth. On the other hand, Asia's population grew the most, with a steady and steep increase. Africa's population also grew significantly. 

```{r q5-task3}
# Filter out Kuwait from the 1952 data
gapminder_1952 <- gapminder_1952 %>% filter(country != "Kuwait")

# Top 5 countries by GDP per capita in 1952 (excluding Kuwait)
top_5 <- gapminder_1952 %>%
  arrange(desc(gdpPercap)) %>%
  slice(1:5)

# Bottom 5 countries by GDP per capita in 1952
bottom_5 <- gapminder_1952 %>%
  arrange(gdpPercap) %>%
  slice(1:5)

# Combine the top and bottom 5 countries
top_5_countries <- top_5 %>% pull(country)
bottom_5_countries <- bottom_5 %>% pull(country)

# Filter the full dataset for top and bottom 5 countries
gapminder_top_5 <- gapminder %>% filter(country %in% top_5_countries)
gapminder_bottom_5 <- gapminder %>% filter(country %in% bottom_5_countries)
gapminder_top_5 <- gapminder_top_5 %>% mutate(group = "Top 5")
gapminder_bottom_5 <- gapminder_bottom_5 %>% mutate(group = "Bottom 5")
combined_data <- bind_rows(gapminder_top_5, gapminder_bottom_5)

ggplot(combined_data, aes(x = year, y = gdpPercap, color = country)) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  facet_wrap(~ group, ncol = 1, scales = "free_y") +  # Stack vertically
  labs(title = "GDP per Capita Trends: Top 5 vs Bottom 5 Countries in 1952 (Excluding Kuwait)",
       x = "Year",
       y = "GDP per Capita",
       color = "Country") +
  theme_minimal() +
  theme(legend.position = "bottom")

```
It seems that all of the countries grew in terms of gdp per capita. I expected to see a downturn for some countries, however, it seems everyone is growing. However, it is important to note that the scale of growth is very different. 

```{r}
# Normalize GDP per capita to 1952 values
combined_data_normalized <- combined_data %>%
  group_by(country) %>%
  mutate(gdpPercap_normalized = gdpPercap / gdpPercap[year == 1952] * 100) %>%
  ungroup()

# Calculate average normalized GDP per capita for top 5 and bottom 5
average_data <- combined_data_normalized %>%
  group_by(group, year) %>%
  summarize(avg_gdpPercap_normalized = mean(gdpPercap_normalized, na.rm = TRUE), .groups = "drop")

# Plot the averages on the same graph
ggplot(average_data, aes(x = year, y = avg_gdpPercap_normalized, color = group)) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  labs(title = "Average Normalized GDP per Capita Trends: Top 5 vs Bottom 5",
       x = "Year",
       y = "Average GDP per Capita (Normalized to 1952 = 100)",
       color = "Group") +
  theme_minimal() +
  theme(legend.position = "bottom")
```
- While I expected more difference, it seems that after normalising, their growth is very similar. The top 5 countries are still growing faster than the bottom 5, especially but the difference is not as stark as I expected. 
